The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-05-13 paper.qmd
fixed up other bits

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
-->

@@

# What is Uncertainty?
In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.  Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

@@

Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy. These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it. The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference. This means that "uncertainty" is not defined by our data, but rather by the *questions we ask about it*. Unfortunately, uncertainty visualisation papers are not free from this issue. Usually papers that are confused on this concept will ask a question that is impossible to answer or does not make sense. Confusion about this core concept is incredibly common and rarely noticed, so much so that this entire paper could simply be a list of papers that have this issue. Instead, we have a handful of highlights.

The first common type of mistake is to present participants with insufficient information to answer a question due to the underlying belief that uncertainty is a latent product of the data. @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. @Hofman2020 performed an experiment where participants were asked to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared despite performing inference on completely different distributions. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference. Both studies highlight a clear belief that uncertainty is related to descriptive statistics, not inferential statistics.

@@

Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem. @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modellers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty. @Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference. While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 4: Organise uncertainty quantification*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship. This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.

# What is Uncertainty Visualisation?
If we simply built the idea of uncertainty visualisation up from "uncertainty" which we already know has a precarious definition we would understand the field of "uncertainty" visualisation must also be precarious. You cannot quantify something that is not defined, and you can't visualise something you cannot quantify. If we cannot quantify the full concept uncertainty in any meaningful way (other than with examples such as missing values or a PDF) then there is no such thing as an "uncertainty" visualisation. Unfortunately this problem is not quite that simple.

## Definitions of "uncertainty visualisation"
Before discussing the issues facing uncertainty visualisation due to a poor concept *of* uncertainty visualisation, we want to establish two ideas that should be considered in conjunction with the lacking definition of uncertainty. The first is the sidestepping of the "uncertainty" definition problem by suggesting "uncertainty visualisation" is inherently different. Some authors may consider "uncertainty" to be poorly defined, but "uncertainty visualisation" to be well established. We will spend the next section explaining why that is not the case by discussing several taxonomies of uncertainty visualisation.

In order to visualise something, it needs to be quantifiable. In order to quantify something, it needs to be mathematically defined. The broad classes of what is considered "uncertainty" rarely have mathematical defintions, and therefore, cannot be meaningfully visualised. For these reasons simple quantifiable uncertainties, such as confidence intervals  dominate the literature, while visualisations for more complicated sources of uncertainty, such as the effects of assumptions, imputed missing variables, and model choices remain elusive.

A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PMF and would already be considered an uncertainty visualisation were our statistic of interest a new oberservation from our data set. These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.

Since the concept of "uncertainty visualisation" is harder to define than "uncertainty", "uncertainty visualisation" taxonomies are typically less thorough than taxonomies for "uncertainty". They also tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects. While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures.

Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution"). This conceptualization could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous. However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty. Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important. (*Cite Gap 9: Other uncertainty vis taxonomies*)

There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
1. Implicit (a sample) or explicit (a depiction of mass)
2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
4. coincidence (uncertainty and data are represented in the same plot) or adjacent
5. static or dynamic (animated, interactive etc)
Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014]. This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution). A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally align with the intrinsic/extrinsic distinction. The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two". This consideration will become more relevant later in this review.

![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy. It can be seen that most of the categories can be considered a change to the typical error bar chart. Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-location width=70%}

Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general. Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation. The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical. @Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualisation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general. The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model. Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.

## Why does the field of "uncertainty visualisation" exist at all?
It is unclear exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.  
(*Cite Gap 10: Reason uncertainty visualisation is different*)
Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 10*). These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.

Uncertainty is noise, but most evaluation experiments measure it as signal. Uncertainty acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020]. Noise and signal are not an inherent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *suppresses* it, if that message is invalid.  When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation. Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisation and the results of the studies in the field. Uncertainty is poorly defined, therefore authors do not understand it is not an inherent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess.

If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables. This does not seem to be the case.

### Information asymmetry in evaluated plots
Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12: Equal information*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. (*Cite Gap 13: examples of what is defined as "information"*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminiscent of our problems with the definition of "uncertainty".

Works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot, and in doing so allow plots to be considered a statistic [@Vanderplas2020]. This statistic that *The Grammar of Graphics* defines fundamentally a descriptive statistic. @fig-pipeline, and contains the set of necessary and sufficient steps that are required to make a graphic, as defined in *The Grammar of Graphics* [@Leland2005]. It can be seen that the grammar assumes we have our data, performs a set of functions on that data, and produces a statistic. The grammar of graphics fundamentally describes the graphic as it is, not as it *would* be were our sample size to increase to infinity. As a matter of fact, the graphics assumption that we start from a data set that has an unchanging sample size is at odds with inference that typically involves assuming an infinitely large sample size. @fig-infintegraph shows that as n approaches infinity, two graphics that are very similar according to the *The Grammar of Graphics* can diverge into different graphics, while two graphics that are very different, can converge into the same graphic. *The Grammar of Graphics* itself does not recognize this issue, as uncertainty is referred to as "semantics" and is ultimately hand waved away. This does not mean describing a graphic using inferential statistics is completely unheard of. Some graphical descriptors, such as Product Plots [@Wickham2011] describe graphics using the underlying distribution, however the framework is not a substitute for *The Grammar of Graphics*. Other authors have noticed difficulties with inferential statistics and and created an addition to the grammar that should make depicting a distribution easier [@Pu2020; @Kay2023]. Unfortunately, this addition falls short as it does not recognize *why* latent distributions are difficult to describe using *The Grammar of Graphics* and carries over many of the same problems.

This does not mean *The Grammar of Graphics* should be abandoned when we visualize uncertainty, rather, these difficulties are something we should keep in mind when

@Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.

If uncertainty is fundamentally a by-product of inference, then the visualisations as described by *The Grammar of Graphics*

Most uncertainty visualisation experiments are written with a seeming indifference to the grammar of Graphics, despite it being necessary to untangle experimental results. The primary reason for this seems to be the absent definition of uncertainty. Uncertainty is fundamentally related to inference and descriptive statistics cannot have uncertainty.

Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. , such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcomes plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

@@

Alternative information consideration within well defined frameworks may also be worthwhile to consider. @wu2023rational tried to eliminate information asymmetry by essentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task. While this solution is interesting, it may have different results depending on who is applying it as they may use different methods or notice different elements when extracting information from a graphic. The mathematically defined concept of sufficient statistics is also scarcely mentioned likely because it is specific to a particular ground truth statistic, however uncertainty *is* specific to a particular statistic, so it could be a useful avenue for understanding when two plots differ in their ability to answer a particular question.

Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation. @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations". They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot). @Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean. They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution. @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations". They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately. (*Cite Gap 14: Examples of mathematically obvious results*).

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.

It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments. Some authors cited uncertainty was special because of its psychological qualities, but we have yet to see any evidence of that here. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maximums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statistic. If we want to visualise the relationship between a temperature estimate and our forecasted variance on that estimate, can we not just use a scatter plot, as we would any other pair of variables? There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially when the core takeaways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation. So, what is it that makes authors believe uncertainty is different?

# Issues with measuring uncertainty as noise, not signal
The previous section highlights a unique and fascinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

>>>>

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- add visually integrable to fig-tax once I work out what it is
- look through for papers that can be used to cite motivation for uncertainty visualisation experiments
-->

@@

# What is uncertainty?
In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.  Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

@@

The belief that uncertainty is a feature of the data, not the variable of interest, leads to several flabbergasting experiments in the uncertainty visualisation space. In these experiments, it is unclear how the participants were supposed to relate what they were shown to the uncertainty distribution the authors had in mind (or if the authors had any distribution in mind at all). For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. In another experiment, @Hofman2020 asked participants to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.

@@

Additionally, if we view uncertainty through the lens of inference then it is obvious that bias and variance both contribute to uncertainty regardless as to which stage of the analysis they appear. Many papers in uncertainty visualisation are not working from this null hypothesis. For example, @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. In another example, @Kale2019 discusses the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values that should be communicated.

Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy. These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it. The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to communicate this complicated by product of inference. This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the confusing set of existing definitions that leads authors to be unaware what they are supposed to be visualising.  Therefore, it is clear the field would greatly benefit from a strict definition of uncertainty. The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.

# What is an uncertainty visualisation?
Given that there is no definition of "uncertainty" readers may wonder how we can define "uncertainty visualisations". This is a reasonable issue given that you cannot quantify something that is not defined, and you can't visualise something you cannot quantify. This section discusses how the problems caused by the absent definition of uncertainty bleed into our understandings of uncertainty visualisation and creates a field that cannot clearly state what is or is not within its domain.

## Definitions of "uncertainty visualisation"
What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors. This completely untethered concept of an "uncertainty visualiation" leads to many user experiments where the visualisations used have a shocking information asymetry.

Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014]. @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. Many visualisation authors ignore this concept and simply compare two visualisations because they both belong to the class of "uncertainty visualisations". @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis. They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot). @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution. @Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two. @Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean. They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. Ideally this problem would be avoided by defining the information in a graphic using *The Grammar of Graphics* since it is a sufficient summary of the information in a graphic and it allows a plot to be considered a statistic [@Vanderplas2020]. The process for summarising the information in a plot using *The Grammar of Graphics* is depicted in @fig-pipeline [@Leland2005]. Unfortunately, *The Grammar of Graphics* is not perfectly suited to summarise the information in an uncertainty visualisation since it assumes that we start from a finite data set. Very often uncertainty is expressed using a distribution or a resampling method that does not have an upper limit on the "sample size". Additionally many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness [@Kale2018; @Newburger2022; @Hofmann2012], an effect that is likely caused by a difference in human perception of mass versus sample. Additionally, *The Grammar of Graphics* behaves strangely when the sample size is adjusted and it the distance between two graphics becomes less consistent. @fig-infintegraph shows how three graphics defined by *The Grammar of Graphics* can change as n approaches infinity. The two graphics that are very similar according to the *The Grammar of Graphics* diverge into different graphics, and two graphics that are very different, converge into the same graphic. This means the statistics visualised in the plot as well as the *The Grammar of Graphics* may not be an adequate to summarise the information in a plot that is being used for uncertainty visualisation. Some authors have noticed difficulties with visualising distributions using `ggplot2` and created extensions that make showing distributions more intuitive [@Pu2020; @Kay2023], however these extensions do not address the core discrepancies with uncertainty visualisation and *The Grammar of Graphics*. This does not mean *The Grammar of Graphics* should be abandoned when we attempt to summaise the information in an uncertainty visualisation, very often visations use the same sample to generate all their plots, so this issue is not always present, however these limitations should be kept in mind when analysing uncertainty visualisations using *The Grammar of Graphics*.

@@

An alternative way authors consider describing an uncertainty visualisation is to use taxonomies. However, these taxonomies tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects. Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy. It can be seen that most of the categories can be considered a change to the typical error bar chart. The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc). Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}

The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important. There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation. These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation. For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.

## Uncertainty visualisations should perform signal supression
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. Visualising a graphic is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different. Why?
